---
title: "Cell Lines"
subtitle: "Simplification of a simple scRNA-seq dataset of 5 cancer cell lines from (Tian et al., 2019)[https://doi.org/10.1038/s41592-019-0425-8]"
author: "Mariia Bilous, Aurelie Gabriel, Leonard Herault and David Gfeller"
output:
  md_document:
    toc: true # table of content true


---

# Suggested course structute:
We first run a **standard scRNA-seq data analysis pipeline** (i.e., data normalization, feature selection, dimensionality reduction, visualization, clustering and differential expression analysis) using the [Seurat](https://satijalab.org/seurat/index.html) framework. Then, we **simplify** the same dataset by computing *metacells* (i.e., grouping transcriptionally highly similar single cells into metacells). For this, we will use a method developed in our group called [SuperCell](https://github.com/GfellerLab/SuperCell). We will also provide some hints on how *metacells* can be computed using alternative approaches including [MetaCell](https://github.com/tanaylab/metacell), [Metacell2.0](https://metacells.readthedocs.io/en/latest/readme.html), and [SEACell](https://github.com/dpeerlab/SEACells). We will then run **'a standard scRNA-seq data analysis pipeline'** adjusted to the metacell data and compare the results obtained at the single-cell and the metacell levels. 

Then, the participant can use the code we provide to build metacells and run some basic analysis for their datasets 

***
**Some options:**

* Alternatively, we can run an adjusted (ie., sample-weighted) pipeline for metacells or/and a standard (ie, Seurat).

* Subsampling at the same graining level -> compare the results to those obtained at the single-cell level

* We could provide an example of how metacell can be used for RNA-velocity (on another dataset)

* How metacells can be used for data integration (if we have nice examples)

* when and if analyzing their own data, make sure to introduce `cell.annotation` and `cell.split.condition` arguments of `SCimplify()` to avoid mixing of annotated cell types / conditions  within metacells. 

***

# TO DO:

* make `supercell_DimPlot()`
* match colors for cell lines and clusters
* 

```{r Load libraries, message=FALSE, warning=FALSE}
# make a data library (cell lines or Zilionis)
library(SuperCell)
library(Seurat)
library(dplyr)
```

```{r Load data}
proj.name    <- 'cell_lines'
data.folder  <- file.path("..", "data", proj.name)

# load single-cell (sc) count matrix and cell metadata 
sc.counts <- readRDS(file.path(data.folder, "sc_counts_filtered.Rds"))
sc.meta   <- readRDS(file.path(data.folder, "sc_meta_filtered.Rds"))

# Make sure metadata and count matrix have the same cells in the same order
if(!identical(rownames(sc.meta), colnames(sc.counts))){
  stop("Metadata (`sc.meta`) does not correspond to the count matrix (`sc.counts`)")
}
```

# Single-cell level
## Standard downstream analysis  
Run a brief analysis at the single-cell level, lets's use the common [Seurat](https://satijalab.org/seurat/index.html) pipeline

<span style="color: green;"> depending on the participants' experience with Seurat, we can either separately run these steps of the pre-processing or 'skip it' with 'one-line' command. For the moment, it is more step-by -step pre-processing, that can be replaced with : </span> 
`sc <-  NormalizeData(sc, verbose=FALSE) %>% FindVariableFeatures(selection.method = "disp", nfeatures = 1000, verbose=FALSE) %>% ScaleData(verbose=FALSE) %>% RunPCA(verbose=FALSE)`

```{r}
set.seed(12345)
sc <- CreateSeuratObject(counts = sc.counts, project = proj.name, meta.data = sc.meta)
sc
```

### Pre-processing
#### Data normalization
```{r Normalize data and compute a set of highly variable genes, message=FALSE, warning=FALSE}
sc <- NormalizeData(sc, verbose=FALSE)

sc <- FindVariableFeatures(
  sc, 
  selection.method = "disp", # "vst" is default
  nfeatures = 1000,
  verbose=FALSE
  )

hvg <- VariableFeatures(sc, verbose=FALSE)

# Plot variable features 
plot1 <- VariableFeaturePlot(sc)
LabelPoints(plot = plot1, points = hvg[1:20], repel = TRUE)
```

#### Scaling and dimensionality redution
```{r message=FALSE, warning=FALSE}
sc <- ScaleData(sc, verbose=FALSE)
sc <- RunPCA(sc, verbose=FALSE)

# Plot PCA (2D representation of scRNA-seq data) colored by cell line
DimPlot(sc, reduction = "pca", group.by = "cell_line")
```

### UMAP (non-linear dimensionality reduction)
```{r UMAP, message=FALSE, warning=FALSE}
sc <- RunUMAP(sc,  dims = 1:10)

# Plot UMAP (2D representation of scRNA-seq data) colored by cell line
DimPlot(sc, reduction = "umap", group.by = "cell_line")
```

### Clustering
```{r message=FALSE, warning=FALSE}

sc <- FindNeighbors(sc, dims = 1:10)
sc <- FindClusters(sc, resolution = 0.05)

# As it is a toy example with well defined cell types (i.e., cell lines), unsupervised clustering fully recapitulates cell line annotation 
table(sc@active.ident, sc$cell_line)
DimPlot(sc, reduction = "umap", group.by = "ident")

```

### Differential expression analysis 
#### Find Markers of cell lines
```{r message=FALSE, warning=FALSE}
# Set idents to cell lines (as clusters are the same as cell lines)
Idents(sc) <- "cell_line"

# Compute upregulated genes in each cell line (versus other cells)
sc.all.markers <-  FindAllMarkers(sc, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25, test.use = "t")
saveRDS(sc.all.markers, file = file.path(data.folder, "output", "sc_all_markers.Rds"))

# Load markers (backup)
#sc.all.markers <- readRDS(file = file.path(data.folder, "output", "sc_all_markers.Rds"))

# Top markers (select top markers of each cell line)
sc.top.markers <- sc.all.markers %>%
   group_by(cluster) %>%
    slice_max(n = 2, order_by = avg_log2FC)

sc.top.markers
```

#### Plot the expression of some found markers 
```{r fig.height=5, fig.width=9}
VlnPlot(sc, features = sc.top.markers$gene[c(seq(1, 9, 2), seq(2, 10, 2))], ncol = 5, pt.size = 0.0)
```


# Data simplification (coarse-graining) -- Construction of *metacells*

Here we compute metacells using our method called [SuperCell](https://github.com/GfellerLab/SuperCell), but equally, metacells can be computed with [Metacell](https://github.com/tanaylab/metacell), [Metacell2.0](https://metacells.readthedocs.io/en/latest/readme.html) or [SEACell](https://github.com/dpeerlab/SEACells) algorithms and we will see some examples below.
```{r Set parameters}
gamma <- 20 # Graining level

# Compute metacells using SuperCell package
MC <- SCimplify(
  X = GetAssayData(sc), # single-cell log-normalized gene expression data
  genes.use = hvg, 
  gamma = gamma,
  n.pc = 10
)

# Compute gene expression of metacells by simply averaging gene expression within each metacell
MC.ge <- supercell_GE(
  ge = GetAssayData(sc),
  groups = MC$membership
)

# Alternatively, counts can be averaged (summed up) followed by a lognormalization step (this approach is used in the MetaCell and SEACell algorithms)
if(0){
  MC.counts <- supercell_GE(
    ge = GetAssayData(sc, slot = "counts"),
    mode = "sum", # summing counts instead of the default averaging
    groups = MC$membership
  )
  
  MC.ge <- Seurat::LogNormalize(MC.counts, verbose = FALSE)
}
```

## Downstream analysis of metacells
There are two options to perform the downstream analysis: 

* **sample-weighted** when we account for a metacell size at each etep of the analysis 
* **standard** when we treat metacells as single-cell and apply a standard pipeline

## Sample-weighted downstream analysis of *metacells*

For the sample-weighted analysis, we use a pipeline avalable with our [SuperCell]() package.

### Pre-processing

#### Transfer metadata (annotate metacell to a certain cell line)
Since the cell line information is available in this dataset, we can annotate metacells to a certain cell line. Each metacell is annotated to the most abundant cell type in it.
This also allows us to compute metacell *purity*, that is defined as a proportion of the most abundant cell type (use `method = "max_proportion"`) or as Shannon entropy (use `method = "entropy"`).

```{r metacell annotation to cell line and their purity}

MC$cell_line <- supercell_assign(
  cluster = sc.meta$cell_line,          # single-cell assigment to cell lines 
  supercell_membership = MC$membership  # single-cell assignment to metacells
)

method_purity <- c("max_proportion", "entropy")[1]
MC.purity <- supercell_purity(
  clusters = sc.meta$cell_line,
  supercell_membership = MC$membership, 
  method = method_purity
)

# Metacell purity distribution
summary(MC.purity)
#hist(MC.purity, main = paste0("Purity of metacells \nin terms of cell line composition (", method_purity,")"))
```
#### Visualize data using  metcall network 
```{r metacell plot metacell network color cell line}
supercell_plot(
  MC$graph.supercells, 
  group = MC$cell_line, 
  seed = 1, 
  alpha = -pi/2,
  main  = "Metacells colored by cell line assignment"
)
```


#### Dimensionality reduction (sample-weighted PCA)
```{r metacell PCA}
MC$PCA <- supercell_prcomp(
  Matrix::t(MC.ge),
  genes.use = MC$genes.use,  # or a new set of HVG can be computed
  supercell_size = MC$supercell_size, # provide this parameter to run sample-weighted version of PCA,
  k = 20
)

# (!) make supercell_DimPlot function instead of UMAP, tSNE etc
MC$PCA_12 <- list("layout" = data.frame(X1 = MC$PCA$x[,1], X2 = MC$PCA$x[,2]))
supercell_plot_UMAP(
  MC, 
  groups = MC$cell_line,
  UMAP_name = "PCA_12", 
  title = paste0("PCA of metacells colored by cell line assignment")
)
```

#### Clustering (sample-weighted hclust)

Sample-weighted clustering computed with the hierarchical clustering, that may accounts for sample weights
```{r metacell clustering}
## compute distance
D              <- dist(MC$PCA$x)

## cluster metacells
MC$clustering    <- supercell_cluster(D = D, k = 5, supercell_size = MC$supercell_size)

# Plot clustering result
supercell_plot_UMAP(
  MC, 
  groups = factor(MC$clustering$clustering),
  UMAP_name = "PCA_12", 
  title = paste0("PCA of metacells colored by metacell clustering")
)
```


## Standard downstream analysis
For the standard downstream analysis, we can use the well-stablished [Seurat]() pipeline
```{r}

```


