---
title: "Metacell integration"
author: "Leonard Herault, Mariia Bilous, Aurelie Gabriel, and David Gfeller"
date: '2022-05-12'
output:
  github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE, message = FALSE)
```

## library loading

```{r}
library(SuperCell)
library(Seurat)
library(SingleCellExperiment)
library(ggplot2)
library(harmony)
library(reshape2)
```
In this tutorial we will make a short reanalysis of a large [single cell transcriptome atlas of COVID19](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7857060/) at the metacell level. The original study gathered 1.4 millions of cells distributed in 284 samples coming from 196 patients. Here we will focus on a subset of 26 fresh PBMC samples.

## Analysis of one sample from covid 19 dataset

We start by analyzing a fresh PBMC sample from an aged patient that deceased from a severe COVID 19.
We will mainly consider  the fist cell annotation to major PBMC cell types coming from the original study ('majorType' column in the metadata).

As in the original study we will discard immunoglobuline, ribosomal protein and mitochondrial genes for the dimension reduction analysis and the metacell construction.

These lists of genes were retrieved from the [genenames website](www.genenames.org). In this tutorial we provide you the whole gene blacklist as an R object.

```{r}
gene_blacklist <- readRDS("../data/gene_blacklist.rds")
```

# Analysis at the single cell level with Seurat

Quality control (gene/cell filtering) was already done in the original study. We go directly to dimension reduction on highly variable genes

```{r}
pbmc <- readRDS("../data/pbmc_COVID19_sce/S-S086-2_sce.rds")

sceRawToSeurat<- function(pbmc,nfeatures = 1500) {
pbmc <- CreateSeuratObject(counts = assay(pbmc),meta.data = data.frame(colData(pbmc)))
pbmc <- FindVariableFeatures(pbmc,nfeatures = nfeatures)
VariableFeatures(pbmc) <- VariableFeatures(pbmc)[!VariableFeatures(pbmc) %in% gene_blacklist]
pbmc <- NormalizeData(pbmc)
return(pbmc)
}

pbmc <- sceRawToSeurat(pbmc)
```

First we can do a Principal Component Analysis (PCA) of this sample with the Seurat package.

```{r}
pbmc <- ScaleData(pbmc)
pbmc <- RunPCA(pbmc, npcs = 20)
ElbowPlot(pbmc) 
PCAPlot(pbmc,group.by = "majorType")
```


We can compute a 2D UMAP on the 20 first components of the PCA for visualization 

```{r}
pbmc <-RunUMAP(pbmc, dims = c(1:20))
UMAPPlot(pbmc, group.by = "majorType")
#UMAPPlot(pbmc, group.by = "celltype") + NoLegend()

```

We make a metacell Seurat object from the single cells of the sample.
We assign metacell to metadata label (`majorType`, `celltype`) with the maximum absolute abundance within metacell thanks to the `supercell_assign*` function.
We compute purity of metacells according to their assignment to the majorType annotation.
```{r}

makeSeuratMC <- function(pbmc,
                         genes = NULL,
                         metaFields = c("majorType", "sampleID", "Age", "Sex", "celltype", "PatientID", "SARS.CoV.2", "Outcome", "datasets", "CoVID.19.severity", "Sample.time"),
                         returnMC = F) {
  
  if(is.null(genes)) {
    genes <- VariableFeatures(pbmc)
  }

MC <- SCimplify(GetAssayData(pbmc,slot = "data"),  # normalized gene expression matrix 
                 n.pc = 20,
                 k.knn = 5, # number of nearest neighbors to build kNN network
                 gamma = 20, # graining level
                 genes.use = genes )# will be the ones used for integration if input is seurat integrated data


MC$purity <- supercell_purity(clusters = pbmc$majorType,
                           supercell_membership = MC$membership)


for (m in metaFields) {
MC[[m]]<- supercell_assign(clusters = pbmc@meta.data[,m], # single-cell assigment to cell lines (clusters)
                                    supercell_membership = MC$membership, # single-cell assignment to super-cells
                                    method = "absolute")
}

GE <- supercell_GE(as.matrix(GetAssayData(pbmc,slot = "data")),groups = MC$membership)

seuratMC <- supercell_2_Seurat(SC.GE = GE,MC,fields = c(metaFields,"purity"))

res <- seuratMC

if (returnMC) {
  res <- list(seuratMC = seuratMC,SC = MC)
}


return(res)
}

supercells <- makeSeuratMC(pbmc,returnMC = T)
```


We use the Seurat workflow to analyse the sample at the metacell level.
Data scaling and PCA computation are weighted according to metacell size with the `supercell_2_Seurat` function we just used. 

```{r}
seuratMC <- supercells$seuratMC
seuratMC <- RunUMAP(seuratMC,dims = c(1:20))
```

We compute the UMAP dimension reduction on the weighted PCA results for visualization. 
```{r}
UMAPPlot(seuratMC,group.by = "majorType")
```

We can use the Seurat workflow without taking into account the metacell weights by restarting from the data scaling step.
This will overwrite the weighted downstream analysis results in the Seurat object.

```{r}
seuratMC <- ScaleData(seuratMC)
seuratMC <- RunPCA(seuratMC, npcs = 20)
seuratMC <- RunUMAP(seuratMC,dims = c(1:10))
UMAPPlot(seuratMC,group.by = "majorType")
```

To be noted that the splits in the CD8 and DC majorType assigned metacells can be explained by the second level of clustering provided in the original study (`celltype` column in the metadata).

```{r}
UMAPPlot(seuratMC,group.by = "celltype") + NoLegend()
```

We can check the distribution of metacell purities and sizes obtained.

```{r}
ggplot(seuratMC@meta.data,aes(x=orig.ident,y=purity)) + geom_boxplot() + 
  ggplot(seuratMC@meta.data,aes(x=orig.ident,y=size)) + geom_boxplot() 
```


# gene correlation analysis

The drop out noise is reduced with metacell analysis. 

```{r}
VlnPlot(pbmc, features = 'nFeature_RNA') 
VlnPlot(seuratMC, feature = 'nFeature_RNA')
```

This results in a better observation of gene-gene expression correlation. To illustrate this we can look for correlations of expression of CXCL8, one cytokine that came out of the original study that might be involved in severe COVID19, and transcription factors that potentially regulate it.

```{r}
gene <- as.matrix(GetAssayData(seuratMC)["CXCL8",])
correlations <- apply(as.matrix(GetAssayData(seuratMC)), 1, function(x){ cor(gene,x) })

tf <- read.table("../data/transcription.factor.activity.GO0003700.symbol.list")

correlations <- correlations[names(correlations) %in% tf$V1]
head(sort(abs(correlations),decreasing = T),n = 10)
```



```{r}
FeatureScatter(object = pbmc, feature1 = "CXCL8", feature2 = 'KLF10', group.by = "majorType")

FeatureScatter(object = seuratMC, feature1 = "CXCL8", feature2 = 'KLF10', group.by = "majorType")

FeatureScatter(object = pbmc, feature1 = "CXCL8", feature2 = 'FOS', group.by = "majorType")

FeatureScatter(object = seuratMC, feature1 = "CXCL8", feature2 = 'FOS', group.by = "majorType")
```
We can also use the `supercell_GeneGenePlot` to have a view of the size of the metacells.
It outputs a weighted correlation according to supercell sizes.

```{r}
supercell_GeneGenePlot(GetAssayData(seuratMC),
                       gene_x = "CXCL8", gene_y = "FOS",
                       supercell_size = seuratMC$size,
                       clusters = seuratMC$majorType)

supercell_GeneGenePlot(GetAssayData(seuratMC),
                       gene_x = "CXCL8", gene_y = "KLF10",
                       supercell_size = seuratMC$size,
                       clusters = seuratMC$majorType)
```
We can also better appreciate the anti-correlation between CCND2 cycling gene and CDKN1A, coding for a cell cycle inhibitor protein.
```{r}
FeatureScatter(object = pbmc, feature1 = "CCND2", feature2 = 'CDKN1A',group.by = "majorType")

supercell_GeneGenePlot(GetAssayData(seuratMC),
                       gene_x = "CCND2",gene_y = "CDKN1A",
                       supercell_size = seuratMC$size,
                       clusters = seuratMC$majorType)
```


# Data integration

We provide you a tutorial for integration at the single cell level [here](Workbook_2_1__COVID19_integration_single_cells.Rmd) if you want to evaluate the time and the memory needed for this analysis at this scale. Note that even with 32GB of memory the integration of all 26 samples is impossible at the single cell level.


# Data integration at the metacell level

*At least 16GB of memory are required to complete this analysis*. 

If you have more than 20GB, you should be able to integrate all the files we provided into `../data/pbmc_COVID19_sce`. (26 data sets, about 200,000 unique cells).

If you have between 16GB and 20GB of memory, you should be able to integrate all the male samples (which are more evenly distributed across the different conditions than the female samples). We provide you with the list of male samples in the file `../data/lowMemFileList.rds` (17 data sets, about 130,000 individual cells).

If you have less than 16GB of memory, you can try to integrate a small subset of 5~6 samples, but the downstream analysis we propose will not be very meaningful.

We will now integrate at the metacell level a subset of 26 samples of fresh PBMCs from the original study gathering around 200 000 cells.

We will use a gamma of 20 (10 000 metacells if you process the 26 data sets).


```{r}
files <- list.files("../data/pbmc_COVID19_sce",full.names = T)

# Uncomment if you have less than 20GB of memory
# files <- readRDS("../data/lowMemFileList.rds")

# Uncomment if you less than 16GB of memory
# files <- files[c(1:6)]

MC.list <- list()
MC.list[["S-S086-2"]] <- seuratMC
for (f in files[which(files != "../data/pbmc_COVID19_sce/S-S086-2_sce.rds")]) {
  smp <- readRDS(f)
  smp <- sceRawToSeurat(smp)
  seuratMC <- makeSeuratMC(smp)
  MC.list[[seuratMC$sampleID[1]]] <- seuratMC
}

```

We select the highly variable genes across samples for integration using the Seurat procedure.

```{r}
features <- SelectIntegrationFeatures(MC.list,nfeatures = 1500)

features <- features[!features %in% gene_blacklist]
```

# Seurat integration workflow

First we will use the Seurat [integration workflow for large datasets]((https://satijalab.org/seurat/articles/integration_large_datasets.html)) using a reference based integration and the RPCA reduction dimension method.

```{r}
nSingleCells <- 0
for (i in 1:length(MC.list)) {
  MC.list[[i]] <- RenameCells(MC.list[[i]],add.cell.id = unique(MC.list[[i]]$sampleID))
  MC.list[[i]] <- ScaleData(MC.list[[i]],features = features)
  MC.list[[i]] <- RunPCA(MC.list[[i]] ,features = features,npcs = 20)
  nSingleCells <- nSingleCells + sum(MC.list[[i]]$size)
}

```

We can print the corresponding number of single cell we are about to integrate at the metacell level
```{r}
nSingleCells
```


The subset of samples come from male and female patients. 
As gender is often a large source of variation at the transcriptomic level, we will use the largest male and female samples as reference.

```{r}
reference <- which(c("S-S086-2","S-M064") %in% names(MC.list)) 



  
anchors <- FindIntegrationAnchors(object.list = MC.list,
                                  reference = reference, 
                                  reduction = "rpca",
                                  anchor.features = features,
                                  dims = 1:20)

integrated <- IntegrateData(anchorset = anchors, dims = 1:20)


```

We can now use the classical Seurat workflow on the integrated object

```{r}
DefaultAssay(integrated) = "integrated"
integrated <- ScaleData(integrated, verbose = T)
integrated <- RunPCA(integrated, verbose = FALSE)
integrated <- RunUMAP(integrated, dims = 1:20)
```

We check that our integration correctly mixed the different samples in the reduced space by preserving differences between the different PBMC types.
We can also check that we don't have a gender effect in the integrated data.

```{r}
UMAPPlot(integrated,group.by = "sampleID")

UMAPPlot(integrated,group.by = "majorType")

UMAPPlot(integrated,group.by = "Sex")

```

Now we will use [Harmony](https://www.nature.com/articles/s41592-019-0619-0), an other integration method. We will use it inside the [Seurat framework](https://htmlpreview.github.io/?https://github.com/satijalab/seurat.wrappers/blob/master/docs/harmony.html).

```{r message=FALSE}
DefaultAssay(integrated) <- "RNA"
integrated <- ScaleData(integrated,features = features)
integrated <- RunPCA(integrated, npcs = 20, verbose = FALSE,features = features)

integrated <- RunHarmony(integrated,c("sampleID","datasets"),theta = c(2.5,1.5))
integrated <- RunUMAP(integrated,reduction = "harmony",
                      reduction.name = "umap.harmony",
                      dims = 1:20)
```


```{r}
DimPlot(integrated,group.by = "sampleID", reduction = "umap.harmony")
DimPlot(integrated,group.by = "majorType", reduction = "umap.harmony")
```

The original study mainly focused on the analysis of variations in the cell type composition according to COVID19 severity (control, mild/moderate, severe) and the time of sampling (control, progression, convalescence).

```{r}
DimPlot(integrated,group.by = "Sample.time", reduction = "umap.harmony") 
DimPlot(integrated,group.by = "CoVID.19.severity", reduction = "umap.harmony")
```

In this tutorial we will make a short analysis of this at the metacell level regarding the first level of annotation (PMBC major type). We can compute the observed versus expected cell number ratio for each PBMC type for the different conditions.
We have to take into account the supercell size for this analysis.

```{r}
majorTypeCounts <- aggregate(integrated$size, by=list(majorType = integrated$majorType,Severity = integrated$CoVID.19.severity,Time = integrated$Sample.time), FUN=sum)


majorTypeCounts$group = paste(majorTypeCounts$Severity,majorTypeCounts$Time,sep = "_")

contingencyTable <- xtabs(x ~ group + majorType,data = majorTypeCounts)

res <- chisq.test(contingencyTable)

Roe <- res$observed/res$expected
```

As in the original study we observe an increase in the Megakaryocytes cell proportion during the progression of severe/critical COVID compare to the other condition.

```{r}
melted_Roe <- melt(Roe, na.rm = TRUE)
# Heatmap
ggplot(data = melted_Roe, aes(majorType,group,  fill = value))+
 geom_tile() + scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
   midpoint = 1, space = "Lab",
   name="Ro/e") +
  theme_minimal()+ 
 theme(axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 12, hjust = 1))+
 coord_fixed()
```
Without the few metacell assigned to macrophages found in the control samples that overwrite the results: 

```{r}
ggplot(data = melted_Roe[melted_Roe$majorType != "Macro",], aes(majorType,group,  fill = value))+
 geom_tile() + scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
   midpoint = 1, space = "Lab",
   name="Ro/e") +
  theme_minimal()+ 
 theme(axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 12, hjust = 1))+
 coord_fixed()
```

We can do a differentially expressed gene analysis to characterize the gene markers of this population that seem to support progression of severe/critical COVID. We perform this analysis at the metacell level (on averaged normalized counts per metacell).

```{r}
Idents(integrated) <- "majorType"
markersMega <- FindMarkers(integrated,ident.1 = "Mega", only.pos = T)
markersMega <- markersMega[markersMega$p_val_adj < 0.05, ]
markersMega[order(markersMega$avg_log2FC, decreasing = T), ]
```

We retrieved the genes coding for inflammatory cytokines highlighted in the original study.

```{r}
cytokines_Mega_ori_study <- c("PDGFA", "TGFB1", "TNFSF4", "PF4V1", "PF4", "PPBP")
cytokines_Mega_ori_study %in% rownames(markersMega)
Idents(integrated) <- "majorType"
VlnPlot(integrated, features = cytokines_Mega_ori_study, pt.size = 0.1)
```

We can compare with the expression of these cytokines in the S-S086-2 sample at the single cell and the metacell level.

```{r}

Idents(MC.list[["S-S086-2"]]) <- "majorType"

pbmc2 <- pbmc[, pbmc$majorType %in% levels(MC.list[["S-S086-2"]])] #to have the same majorType in MC and single cell object see below
pbmc2$majorType <- factor(pbmc2$majorType,levels = levels(MC.list[["S-S086-2"]]))
Idents(pbmc2) <- "majorType"

VlnPlot(pbmc2, features = cytokines_Mega_ori_study, pt.size = 0.1)
VlnPlot(MC.list[["S-S086-2"]], features = cytokines_Mega_ori_study, pt.size = 0.1)
```
## About metacell purity

We can see that for this PBMC sample the 4 single cell annotated as Neutrophils in the original study were merged with Monocytes at the metacell level.


```{r}
table(pbmc$majorType)
table(seuratMC$majorType)
Neu <- colnames(pbmc[,pbmc$majorType == "Neu"])
Neu
supercells$seuratMC$majorType[supercells$SC$membership[Neu]]
```

To completely match the original study results one possible strategy would be to make metacells per PBMC type 
This will result in completely pure metacells according to the original annotation. 


```{r}

MC_majorTypes <- lapply(X = unique(supercells$seuratMC$majorType), FUN = function(x){SC <- SCimplify(GetAssayData(pbmc[,pbmc$majorType == x],slot = "data"),
       n.pc = 20,
       k.knn = 5, # number of nearest neighbors to build kNN network
       gamma = 20 # graining level
);SC$majorType = rep(x,SC$N.SC);return(SC)})



Neu_MC <- SCimplify(GetAssayData(pbmc[,pbmc$majorType == "Neu"],slot = "data"),
       n.pc = 3,
       k.knn = 4, # number of nearest neighbors to build kNN network
       gamma = 1 # graining level
)

Neu_MC$majorType <- rep("Neu",Neu_MC$N.SC)

MC_majorTypes <- c(MC_majorTypes,list(Neu_MC))

MC_majorTypes2 <- supercell_merge(MC_majorTypes, fields = "majorType")

remove(MC_majorTypes)
gc()

MC_majorTypes_GE <- supercell_GE(as.matrix(GetAssayData(pbmc)[,names(MC_majorTypes2$membership)]),groups = MC_majorTypes2$membership)

MC_majorTypes2$genes.use <- VariableFeatures(pbmc)

seuratPureMC <- supercell_2_Seurat(SC.GE = MC_majorTypes_GE,
                                   SC = MC_majorTypes2,
                                   fields = c("majorType"))
```


We use the Seurat workflow to analyse the sample at the metacell level.
Data scaling and PCA computation are weighted according to metacell size with the `supercell_2_Seurat` function we just used. 

```{r}
seuratPureMC <- RunUMAP(seuratPureMC,dims = c(1:20))
```

We compute the UMAP dimension reduction on the weighted PCA results for visualization. 
```{r}
UMAPPlot(seuratPureMC,group.by = "majorType")
ggplot(seuratPureMC@meta.data,aes(x=orig.ident,y=size)) + geom_boxplot() 
ncol(seuratPureMC)
```

We can use the Seurat workflow without taking into account the metacell weights by restarting from the data scaling step.
This will overwrite the weighted downstream analysis results in the Seurat object.

```{r}
seuratPureMC <- ScaleData(seuratPureMC)
seuratPureMC <- RunPCA(seuratPureMC, npcs = 20)
seuratPureMC <- RunUMAP(seuratPureMC,dims = c(1:20))
UMAPPlot(seuratPureMC,group.by = "majorType")

FeaturePlot(seuratPureMC,features = "nFeature_RNA")

```




```{r}
# sampleList <- str_split_fixed(files,pattern = "/",n=4)[,4]
# sampleList <- str_split_fixed(sampleList,pattern = "_",n=2)[,1]
# 
# sampleLowMem <- unique(integrated$sampleID[integrated$Sex == "M" ])
# sum(integrated$size[integrated$sampleID %in% sampleLowMem])
# 
# sampleLowMemFileList <- files[sampleList %in% sampleLowMem]
# 
# saveRDS(sampleLowMemFileList,"../data/lowMemFileList.rds")
# 
# sizeAll <- aggregate(integrated@meta.data[,"size"],
#                       by = list(sampleID = integrated@meta.data[,"sampleID"]),FUN = sum)
# 
# sizeAll[order(sizeAll$x),]
# 
# sizeF <- aggregate(integrated@meta.data[integrated$Sex == "F","size"],
#                       by = list(sampleID = integrated@meta.data[integrated$Sex == "F","sampleID"]),FUN = sum)
# 
# largestF <- sizeF$sampleID[which.max(sizeF$x)]
# 
# sizeH <- aggregate(integrated@meta.data[integrated$Sex == "M","size"],
#                       by = list(sampleID = integrated@meta.data[integrated$Sex == "M","sampleID"]),FUN = sum)
# 
# largestH <- sizeH$sampleID[which.max(sizeH$x)]
```
