---
title: "tutorial_single_cell_integration"
author: "Leonard Herault, Mariia Bilous, Aurelie Gabriel, and David Gfeller"
date: '2022-06-07'
output: html_document
---
---
title: "Metacell integration"
author: "Leonard Herault"
date: '2022-05-12'
output:
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE, message = FALSE)
```

## library loading

```{r}
library(SuperCell)
library(Seurat)
library(SingleCellExperiment)
library(ggplot2)
library(harmony)
library(reshape2)



```


# Data integration at the single cell level

If you have 30 GB of memory you can try to integrate `nSmp = 13` samples.
If you have 15 GB of memory you can try to integrate `nSmp = 7` samples.

```{r}

nSmp = 13

referenceFileIdx <- c(13,26)

sampleFileIdx <- c(1:nSmp)[-c(13,25)]

sampleFileIdx <- c(referenceFileIdx,sampleFileIdx[1:(nSmp-2)])

start_time <- Sys.time()

files <- list.files("../data/pbmc_COVID19_sce",full.names = T)[sampleFileIdx]

gene_blacklist <- readRDS("../data/gene_blacklist.rds")




sceRawToSeurat<- function(pbmc,nfeatures = 1500) {
pbmc <- CreateSeuratObject(counts = assay(pbmc),meta.data = data.frame(colData(pbmc)))
pbmc <- FindVariableFeatures(pbmc,nfeatures = nfeatures)
VariableFeatures(pbmc) <- VariableFeatures(pbmc)[!VariableFeatures(pbmc) %in% gene_blacklist]
pbmc <- NormalizeData(pbmc)
return(pbmc)
}

sc.list <- list()


for (f in files) {
  print(f)
  smp <- readRDS(f)
  smp <- sceRawToSeurat(smp)
  sc.list[[as.character(unique(smp$sampleID))]] <- smp
}

reference <- which(c("S-S086-2","S-M064") %in% names(sc.list)) 


features <- SelectIntegrationFeatures(sc.list,nfeatures = 1500)

features <- features[!features %in% gene_blacklist]

for (i in 1:length(sc.list)) {
  sc.list[[i]] <- RenameCells(sc.list[[i]],add.cell.id = unique(sc.list[[i]]$sampleID))
  sc.list[[i]] <- ScaleData(sc.list[[i]],features = features)
  sc.list[[i]] <- RunPCA(sc.list[[i]] ,features = features,npcs = 20)
}



anchors <- FindIntegrationAnchors(object.list = sc.list,
                                  reference = reference, 
                                  reduction = "rpca",
                                  anchor.features = features,
                                  dims = 1:20)

integrated <- IntegrateData(anchorset = anchors, dims = 1:20)

end_time <- Sys.time()

end_time - start_time 

gc()

```

# Classical single cell analysis

We can now use the Seurat workflow on the integrated object

```{r}
DefaultAssay(integrated) = "integrated"
integrated <- ScaleData(integrated, verbose = T)
integrated <- RunPCA(integrated, verbose = FALSE)
integrated <- RunUMAP(integrated, dims = 1:20)
```


We check that our integration correctly mixed the different samples in the reduced space by preserving differences between the different PBMC types.
We can also check that we don't have a gender effect in the integrated data.

```{r}
UMAPPlot(integrated,group.by = "sampleID")

UMAPPlot(integrated,group.by = "majorType")

UMAPPlot(integrated,group.by = "Sex")

```
```{r}
ncol(integrated)
```

